### 2021-03-02 paper/main.tex

% to-do
% -----
% - Is the GP relationship fully encoded in Bishop? If so, cite; if not, get more precise about what's new here. Also, maybe add a footnote in which we give the translation between our notation and Bishop's notation. Oh and thank Espinoza.
% - Should we mention the "dual" representation? HOGG thinks we are going to the dual when we go from XTX to XXT.
% - Should we do a CVMSE vs the s parameter to show that you can choose kernel parameters by CV as well? Let's start by performing an experiment in the notebook. Not for now.
% - To be submitted to EPJ Data Science https://epjdatascience.springeropen.com/about

@@

% Hogg's issues
\setlength{\topmargin}{-0.50in}
\setlength{\headsep}{4ex}
\setlength{\textheight}{9.30in}
\setlength{\oddsidemargin}{0.5in}
\setlength{\textwidth}{5.50in}
\setlength{\parindent}{0.18in}
\pagestyle{myheadings}
\markright{\textcolor{gray}{\textsf{\upshape Fitting very flexible models / Hogg \& Villar}}}
\newlength{\figurewidth}
\setlength{\figurewidth}{0.94\textwidth}
\setlength{\textfloatsep}{1ex}
\setlength{\floatsep}{1ex}
\frenchspacing\sloppy\sloppypar\raggedbottom

@@

% text definitions
\newcommand{\documentname}{\textsl{Review}}
\newcommand{\sectionname}{Section}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\code}[1]{\texttt{#1}}

@@

\noindent
\textbf{David W. Hogg}\footnote{Corresponding author: David W. Hogg \href{mailto:david.hogg@nyu.edu}{david.hogg@nyu.edu}.} \\
\textsl{\footnotesize Flatiron Institute, a division of the Simons Foundation \\
Center for Cosmology and Particle Physics, Department of Physics, New York University \\
Center for Data Science, New York University \\
Max-Planck-Institut f\"ur Astronomie, Heidelberg}

@@

\paragraph{Abstract:} There are many uses for linear fitting; the context here is interpolation and denoising of data, as when you have calibration data and you want to fit a smooth, flexible function to those data.
Or you want to fit a flexible function to de-trend a time series or normalize a spectrum.
In these contexts, investigators often choose a polynomial basis, or a Fourier basis, or wavelets, or something equally general.
They also choose an order, or number of basis functions to fit, and (often) some kind of regularization.
We discuss how this basis-function fitting is done, with ordinary least squares and extensions thereof.
We emphasize that it is often valuable to choose \emph{far more parameters than data points}, despite folk rules to the contrary:
Suitably regularized models with enormous numbers of parameters generalize well and make good predictions for held-out data; over-fitting is not (mainly) a problem of having too many parameters.
It is even possible to take the limit of infinite parameters, at which, if the basis and regularization are chosen correctly, the least-squares fit becomes the mean of a Gaussian process.
We recommend cross-validation as a good empirical method for model selection (for example, setting the number of parameters and the form of the regularization),
and jackknife resampling as a good empirical method for estimating the uncertainties of the predictions made by the model.
We also give advice for building stable computational implementations.

@@

If you have encountered and solved problems like these, you have made these three kinds of choices (sometimes implicitly).
The second choice---about the number of coefficients to fit---is usually made heuristically, and often subject to the strongly believed opinion that you \emph{must have fewer parameters than data points}.
Here we are going to show that this folk rule is not valid; you can go to extremely large numbers of coefficients without trouble.
But like most folk rules, it has a strong basis in reality: There are extremely bad choices possible for the number of coefficients, and especially when the number of parameters is close or comparable to the number of data points.
As we will discuss below, these choices ought to be made with care.

In many cases the third kind of choice---about regularization---is made implicitly, not explicitly.
Here we are going to emphasize this choice and its importance, and its value in improving your results.

@@

An important assumption or setting for the problems we are addressing in this \documentname\ will be that you care about predicting new data or interpolating the data, but you explicitly \emph{don't care} about the parameters of the fit or the weights of the basis functions \foreign{per se}.
In this setting there are no important meanings to the components of the model.
That is---for us---only the data exist.
The details of the model are just choices that permit high-quality interpolations and predictions in the space of the data.

@@

In some contexts you have strong beliefs about the noise affecting your measurements. In other cases you don't.
In some cases you have strong reasons to use a particular basis.
In other cases you don't.
The differences in these beliefs and the differences in your objectives will change what methods you choose and how you use and analyze them.
We'll try to be useful to you no matter where you're at.
This \documentname\ delivers a theoretical and practical discussion of linear regression and function interpolation; it does not deliver new research results.
Our only novel contribution is to make a specific connection between regularized linear regression and Gaussian processes.

@@

We are going to fit these data $y_i$ with a linear sum of $p$ basis functions $g_j(t)$. These basis functions are functions of the coordinates $t$. That is, our implicit generative model is
\begin{equation}
    y_i = \sum_{j=1}^p \beta_j\,g_j(t_i) + \mathrm{noise}
    ~,
\label{eq.model}
\end{equation}
where the $p$ values $\beta_j$ are parameters or coefficients of the linear fit. We can assemble the evaluations of the $p$ functions $g_j(t)$ at the $n$ data coordinates $t_i$ into a $n\times p$ design matrix or feature matrix $X$ such that
\begin{equation}
    [X]_{ij} = g_j(t_i)
    ~.
\end{equation}
The transformation of the $n$ locations $t_i$ into a $n\times p$ matrix $X$ is the opposite of a dimensionality reduction.
It is called variously a \emph{feature map} or a \emph{feature embedding}.
We like ``embedding'' because it (almost always) raises the dimensionality of the locations $t$ into the $p$-dimensional space of the rows of $X$.

@@

But recall our setting:
We are using the linear fit to interpolate the data, or de-noise the data, or predict new data.
In these contexts, we don't care about the parameter vector $\hat{\beta}$ itself.
We care only about the predictions at a set of new ``test'' locations $t_\ast$, which will usually be different from the training locations $t_j$.
From the new test times $t_\ast$ we create the test feature matrix $X_\ast$ (by the same feature embedding functions $g_j(t)$).
The prediction $\hat{Y}_\ast$ for the $y$ values at the test locations $t_\ast$ becomes
\begin{equation}\label{eq:OLS}
    \hat{Y}_\ast = X_\ast\,(X^\top X)^{-1}\,X^\top Y
    ~.
\end{equation}
Examples of OLS applied to some toy data\footnote{The algorithm by which the toy data were made---and indeed all of the code used to make the figures for this \documentname---is available online at \url{https://github.com/davidwhogg/FlexibleLinearModels}.} are shown in \figurename~\ref{fig:ols1}.
This form \eqref{eq:OLS} of the prediction of new test data is called ordinary least squares (OLS). It has many good properties, some of which are encoded in the Gauss--Markov theorem.
In particular, if the noise contributions in the model given in equation \eqref{eq.model} are uncorrelated, have zero mean, and equal variances for $i=1,\ldots, n$ then the Gauss--Markov theorem states that the OLS estimator has the lowest variance within the class of unbiased estimators that are linear in $Y$ (see, for example, \citealt{esl}, Ch.~3).
\begin{figure}[t]
    \begin{mdframed}
    \includegraphics[width=\figurewidth]{./OLS-under.pdf}
    \caption{Ordinary least-squares (OLS) fits (continuous lines) to a set of example data points (black dots). Fits are shown for different values of the number of basis functions $p$; there are $n=23$ data points. Here (and in all the \figurename s to follow) we are using the Fourier basis functions in \eqref{eq:basis} and shown in \figurename~\ref{fig:basis}. The data $y_i$ were generated using a function that does not reside in the function space spanned by the basis. Here the $X_\ast$ matrices used in the predictions $\hat{Y}_\ast$ were generated from a fine grid of locations in the location coordinate $t$; the plots of the fine grid of predictions are the continuous lines. The fits with larger $p$ have more flexibility to fit the data than the fit with $p=3$, but the fit at the highest $p$ shows evidence of over-fitting.}
    \label{fig:ols1}
    \end{mdframed}
\end{figure}

@@

Although the OLS prediction \eqref{eq:OLS} is affine invariant with respect to $p$-dimensional transformations, it is \emph{not} affine invariant with respect to $n$-dimensional transformations, such as a re-weighting of the input data. Indeed, if you know weights or inverse variances for your data points, conceptually you can put them into a weight matrix $C^{-1}$ (written this way to emphasize that reweighting is usually inverse-variance weighting) and write
\begin{equation}\label{eq:wls}
    \hat{Y}_\ast = X_\ast\,(X^\top C^{-1}\,X)^{-1}\,X^\top C^{-1}\,Y
    ~,
\end{equation}
where the weight matrix $C^{-1}$ is $n\times n$ (and often diagonal in standard applications). The weight matrix $C^{-1}$ is sometimes called the information tensor (or information matrix) and its inverse $C$ is often called the covariance matrix or the noise variance tensor.

This form \eqref{eq:wls} of least squares is called \emph{weighted least squares} (WLS) because of the data weighting (not to be confused with \emph{feature} weighting, to appear below).
It is also called \emph{chi-squared fitting} because it optimizes the scalar objective commonly called chi-squared:
\begin{equation}
    \chi^2 = (Y - X\,\beta)^\top C^{-1}\,(Y - X\,\beta)
    ~.
\end{equation}
(If that isn't obviously chi-squared to you, recall that $C^{-1}$ is often diagonal and has the inverses of the squares of the data uncertainties on that diagonal.)
In \figurename~\ref{fig:wls} a comparison of OLS and WLS is shown, for a case of non-trivial data weights, where the data weights are set to be the inverse squares of individual data-point uncertainties.
\begin{figure}[t]
    \begin{mdframed}
    \includegraphics[width=\figurewidth]{./WLS.pdf}
    \caption{Comparison of ordinary least-squares (OLS) and weighted least-squares (WLS) fits (continuous lines) to the example data (black dots). In order to illustrate the differences, we assigned non-trivial error bars to the data. The error bars are ignored in the OLS fit, but in the WLS fit, the weight matrix $C^{-1}$ is diagonal with the diagonal entries set to the inverses of the squares of those error bars. The WLS fit ``pays less attention to'' the points on the left with the largest error bars.}
    \label{fig:wls}
    \end{mdframed}
\end{figure}

We will say more about noisy data and the propagation of uncertainties below in \sectionname~\ref{sec:uncertainty}.
It might be crossing your mind that there are uncertainties not just in the data points $y_i$, but also often in the \emph{locations} $t_i$ of the data as well. It turns out that taking the latter into account is a much harder problem; we will discuss this briefly in \sectionname~\ref{sec:uncertainty}.

@@

The ridge brings with it a choice: How to set the hyper-parameter $\lambda$?
We generally recommend cross-validation, to be discussed below in \sectionname~\ref{sec:dd}.
Ridge regression is not affine invariant in the sense that the standard OLS prediction \eqref{eq:OLS} is; that is, the effect of the regularization depends on the amplitudes and linear combinations of features placed in the feature matrix.
This will become important later when we consider feature weights below.
It is also the case that the regularization need not be proportional to the identity matrix: In principle any positive definite matrix $\Lambda$ could be used in place of $\lambda\,I$; this makes sense to consider when you have detailed prior beliefs about all the parameters $\beta_j$ or if those parameters are measured with different units (say).

You can combine both the point weighting from WLS and the generalized ridge regression into a weighted ridge that looks like
\begin{equation}\label{eq:foo}
    \hat{Y}_\ast = X_\ast\,(X^\top C^{-1}\,X + \Lambda)^{-1}\,X^\top C^{-1}\,Y
    ~.
\end{equation}
This form has good properties for many real-world physics applications, where data-point error bars are often known, and functions are often expected to be smooth.
We'll discuss this form \eqref{eq:foo} more below in \sectionname~\ref{sec:fwols},
but briefly we can say here that it appears in Bayesian inference contexts where the data points $y_i$ are treated as having Gaussian noise associated with them (with known variance tensor or covariance matrix $C$) and there is a Gaussian prior on the parameter vector $\beta$ (with known variance $\Lambda^{-1}$).
We have discussed that model elsewhere (\citealt{products}).\footnote{In our previous discussion of this product of Gaussians, the notation differs. What's called $\Lambda^{-1}$ here is called $\Lambda$ in \citet{products}.}
It is also useful sometimes to think of the ``units'' or dimensions of the quantities in \eqref{eq:foo}: In the Bayesian setting, the units of $C$ would be the square of the units of $Y$ and the units of $\Lambda^{-1}$ would be the square of the units of the ratio $Y/X$ (the square of the units of $\beta$).

@@

We are taught folklore, at a young age, that we can never fit for more parameters than we have data. That is, we can never work at $p>n$.
This isn't true!
Not only is it the case that we \emph{can} work at $p>n$, in many cases we \emph{should} work at $p>n$, and many real-world regressions do.
But it is true that the $p>n$ regime is indeed strange!
In the over-parameterized case, there are typically many settings of the parameters $\beta_j$ that will literally zero out the differences between the data $Y$ and the linear prediction $X\,\beta$.
The OLS solution, in this case, is defined (somewhat arbitrarily) to be the minimum-norm parameter vector $\beta$ that interpolates the data:
\begin{equation}\label{eq:opt2}
    \hat{\beta} = \arg\min_\beta \|\beta\|_2^2 ~~\mbox{subject to}~~ Y = X\,\beta
    ~.
\end{equation}
Technically this formulation depends on an additional assumption that the feature matrix $X$ is full rank or that the data $Y$ lie in the subspace spanned by $X$.
This is true almost always when $p>n$.
This optimization is again convex and has a unique solution, although that solution will depend on feature weights that we discuss in a moment.
When the investigator knows uncertainties on the training data points $y_i$, this expression will change a bit; we return to that in the next \sectionname.
The under-parameterized and over-parameterized optimization statements \eqref{eq:opt1} and \eqref{eq:opt2} can be unified into one form by considering the limit of light L2 regularization:
\begin{equation}\label{eq:opt3}
    \hat{\beta} = \lim_{\lambda\to 0^+}\left[\arg\min_\beta \|Y - X\,\beta\|_2^2 + \lambda\,\|\beta\|_2^2\right]
    ~;
\end{equation}
in the limit, this delivers the OLS solution in either case and doesn't require the constraint in \eqref{eq:opt2} to be satisfied.

In the over-parameterized case ($p>n$), the prediction looks like
\begin{equation}\label{eq:OLS2}
    \hat{Y}_\ast = X_\ast\,X^\top (X\,X^\top)^{-1}\,Y
    ~,
\end{equation}
provided that $X\,X^\top$ is invertible (which will usually be the case).
Like \eqref{eq:OLS}, this prediction is \emph{also} called ordinary least squares (OLS), or sometimes ``min-norm least-squares'' to emphasize the point that it is making the minimum-norm choice of $\beta$ among many degenerate solutions.
Examples of OLS fits in the over-parameterized regime are shown in \figurename~\ref{fig:ols2}.
Note that the fits with different numbers of parameters $p$ lead to very different predictions, but they all go through the data exactly.
\begin{figure}[t]
    \begin{mdframed}
    \includegraphics[width=\figurewidth]{./OLS-over.pdf}
    \caption{Ordinary least-squares (OLS) or min-norm least-squares fits (continuous lines) to a set of example data points (black dots), but now for a few over-parameterized cases. There are $n=23$ data points. As in \figurename~\ref{fig:ols1}, the data points $y_i$ were generated using a function that does not precisely reside in the function space spanned by the basis. The three fits are very different, but they all go through all the data points exactly. As $p$ gets large, the fit function approaches $y=0$ almost everywhere; this is a consequence of Plancharel's Theorem (see text).}
    \label{fig:ols2}
    \end{mdframed}
\end{figure}

It is slightly off-topic to notice that in \figurename~\ref{fig:ols2}, as $p$ gets very large, the OLS solution approaches $y=0$ everywhere that it can.
This behavior is a direct consequence of Plancharel's theorem, which states that the Fourier transform is unitary (see for instance \citealt{folland2016course}).
This means that the integral over frequency of the square of the Fourier transform is equal to the integral over location of the square of the original function.
Thus the min-norm solution delivered by OLS, which chooses the interpolating function that minimizes the squares of the component amplitudes $\beta$ in the Fourier basis used to make \figurename~\ref{fig:ols2}, will choose the interpolating function $y(t)$ that minimizes the mean square of the value of the function in the location space.
It will try to stay as close to $y=0$ as possible.
That's probably not desirable in most applications! We will fix that problem in the next \sectionname.

@@

In particular, the CVMSE for the feature-weighted OLS fits is bad when $p\approx n$, and much better at $p\ll n$ and $p \gg n$.
This phenomenology is not particular to this problem.
It is extremely general.
There is an effect known in many kinds of regression called ``double descent'' or ``peaking phenomenon'' or ``jamming'' in which predictive accuracy becomes very poor when the number of free parameters $p$ comes close to the number of data points $n$ (\citealt{jain198239}, \citealt{spigler2018jamming}, \citealt{geiger2019jamming}). For linear models, the ``risk''---the out-of-sample prediction error---blows up when the model capacity just becomes excessive at $p=n$ (\citealt{hastie2019surprises}).
The fundamental reason for this phenomenon is that the ordinary least-squares estimates \eqref{eq:OLS} and \eqref{eq:OLS2} require computation of the inverses $(X^\top X)^{-1}$ and $(X\,X^\top)^{-1}$ respectively, which are very badly conditioned around $p\approx n$.
This translates to a large variance for the estimator $\hat \beta$ which implies a large risk.
One way to think about this is that when the condition number of the matrix $X^\top X$ or $X\,X^\top$ is large, some directions in the data space---some linear combinations of the elements of the data vector $Y$---are very strongly amplified.\footnote{The condition number comes back up below in \sectionname~\ref{sec:implementation}. It turns out that when the condition number is large, not only do the output predictions become very sensitive to the input training data, but also the numerical (computational) stability of the linear algebra can also be badly affected. But we emphasize here that the risk goes bad when the condition number is large \emph{even if} it is possible to perform the linear algebra correctly at high precision.}
This makes the regression unreasonably sensitive to noise in the data.
The high-risk behavior at $p\approx n$ disappears with certain kinds of regularization.
These include the ridge regularization shown in \figurename~\ref{fig:cv} (also known as Tikhonov regression), but also early stopping (\citealt{hastie2019surprises}), and dimensionality reduction (\citealt{huang2020dimensionality}).

>>>>

%Changes since submission on  2021-03-02
%--
%Do it!

% to-do
% -----
% - HOGG: Finish going through Acquaviva comments and incorporate / assess.
% - BOTH: Take one more look through the abstract (and title) and see what needs to be adjusted.
% - HOGG. It is a kernel even when p \ne \infty! Re-word stuff in section 7.
% - Should we do a CVMSE vs the s parameter to show that you can choose kernel parameters by CV as well? Let's start by performing an experiment in the notebook. Not for now.

@@

% Hogg's issues
\setlength{\topmargin}{-0.50in}
\setlength{\headsep}{4ex}
\setlength{\textheight}{9.30in}
\setlength{\oddsidemargin}{0.5in}
\setlength{\textwidth}{5.50in}
\setlength{\parindent}{0.18in}
\pagestyle{myheadings}
\markright{\textcolor{gray}{\textsf{\upshape Fitting very flexible models}}}
\newlength{\figurewidth}
\setlength{\figurewidth}{0.94\textwidth}
\setlength{\textfloatsep}{1ex}
\setlength{\floatsep}{1ex}
\frenchspacing\sloppy\sloppypar\raggedbottom

@@

% text definitions
\newcommand{\documentname}{\textsl{expository note}}
\newcommand{\sectionname}{Section}
\newcommand{\foreign}[1]{\textsl{#1}}
\newcommand{\code}[1]{\texttt{#1}}

@@

\noindent
\textbf{David W. Hogg} \\%\footnote{Corresponding author: David W. Hogg \href{mailto:david.hogg@nyu.edu}{david.hogg@nyu.edu}.} \\
\textsl{\footnotesize Flatiron Institute, a division of the Simons Foundation \\
Center for Cosmology and Particle Physics, Department of Physics, New York University \\
Center for Data Science, New York University \\
Max-Planck-Institut f\"ur Astronomie, Heidelberg}

@@

\paragraph{Abstract:} There are many uses for linear fitting; we consider here the interpolation and denoising of data, as when the goal is to fit a smooth, flexible function to a set of noisy data points.
Investigators often choose a polynomial basis, or a Fourier basis, or wavelets, or something equally general.
They also choose an order, or number of basis functions to fit, and (often) some kind of regularization.
We discuss how this basis-function fitting is done, with ordinary least squares and extensions thereof.
We emphasize that it can be valuable to choose \emph{far more parameters than data points}, despite folk rules to the contrary:
Suitably regularized models with enormous numbers of parameters generalize well and make good predictions for held-out data; over-fitting is not (mainly) a problem of having too many parameters.
It is even possible to take the limit of infinite parameters, at which, if the basis and regularization are chosen correctly, the least-squares fit becomes the mean of a Gaussian process, or a kernel regression.
We recommend cross-validation as a good empirical method for model selection (for example, setting the number of parameters and the form of the regularization),
and jackknife resampling as a good empirical method for estimating the uncertainties of the predictions made by the model.
We also give advice for building stable computational implementations.

@@

If you have encountered and solved problems like these, you have made these three kinds of choices (sometimes implicitly).
The second choice---about the number of coefficients to fit---is usually made heuristically, and often subject to the strongly believed opinion that we \emph{must have fewer parameters than data points}.
Here we show that this folk rule is not valid; we can go to extremely large numbers of coefficients without trouble.
But like most folk rules, it has a strong basis in reality: There are extremely bad choices possible for the number of coefficients, and especially when the number of parameters is close or comparable to the number of data points.
As we will discuss below, these choices ought to be made with care.

In many cases the third kind of choice---about regularization---is made implicitly, not explicitly.
Here we are going to emphasize this choice and its importance, and its value in improving results.

@@

An important assumption or setting for the problems we are addressing in this \documentname\ will be that we care about predicting new data or interpolating the data, but we explicitly \emph{don't care} about the parameters of the fit or the weights of the basis functions \foreign{per se}.
In this setting there are no important meanings to the components of the model.
That is---for us---only the data exist.
The details of the model are just choices that permit high-quality interpolations and predictions in the space of the data.

@@

In some contexts we have strong beliefs about the noise affecting our measurements. In other cases we don't.
In some cases we have strong reasons to use a particular basis.
In other cases we don't.
The differences in these beliefs and the differences in our objectives will change what methods we choose and how we use and analyze them.

This \documentname\ delivers a theoretical and practical discussion of linear regression and function interpolation; it does not deliver new research results. Our only novel contribution is to make an explicit connection between regularized linear regression and Gaussian processes.

@@

We are going to fit these data $y_i$ with a linear sum of $p$ basis functions $g_j(t)$. These basis functions are functions of the coordinates $t$. That is, our implicit generative model is
\begin{equation}
    y_i = \sum_{j=1}^p \beta_j\,g_j(t_i) + \mathrm{noise}
    ~,
\label{eq.model}
\end{equation}
where the $p$ values $\beta_j$ are parameters or coefficients of the linear fit. We can assemble the evaluations of the $p$ functions $g_j(t)$ at the $n$ data coordinates $t_i$ into a $n\times p$ design matrix or feature matrix $X$ such that
\begin{equation}
    [X]_{ij} = g_j(t_i)
    ~.
\end{equation}
The transformation of the $n$ locations $t_i$ into a $n\times p$ matrix $X$ is the opposite of a dimensionality reduction.
It is called variously a \emph{feature map} or a \emph{feature embedding}.
The embedding raises the dimensionality of the locations $t$ into the $p$-dimensional space of the rows of $X$.

@@

But recall our setting:
We are using the linear fit to interpolate the data, or de-noise the data, or predict new data.
In these contexts, we typically don't care about the parameter vector $\hat{\beta}$ itself.
We care only about the predictions at a set of new ``test'' locations $t_\ast$, which will usually be different from the training locations $t_j$.
From the new test times $t_\ast$ we create the test feature matrix $X_\ast$ (by the same feature embedding functions $g_j(t)$).
The prediction $\hat{Y}_\ast$ for the $y$ values at the test locations $t_\ast$ becomes
\begin{equation}\label{eq:OLS}
    \hat{Y}_\ast = X_\ast\,(X^\top X)^{-1}\,X^\top Y
    ~.
\end{equation}
Examples of OLS applied to some toy data\footnote{The algorithm by which the toy data were made---and indeed all of the code used to make the figures for this \documentname---is available online at \url{https://github.com/davidwhogg/FlexibleLinearModels}.} are shown in \figurename~\ref{fig:ols1}.
This form \eqref{eq:OLS} of the prediction of new test data is called ordinary least squares (OLS). It has many good properties, some of which are encoded in the Gauss--Markov theorem.
In particular, if the noise contributions in the model given in equation \eqref{eq.model} are uncorrelated, have zero mean, and equal variances for $i=1,\ldots, n$ then the Gauss--Markov theorem states that the OLS estimator has the lowest variance within the class of unbiased estimators that are linear in $Y$ (see, for example, \citealt{esl}, Ch.~3).
\begin{figure}[t]
    \begin{mdframed}
    \includegraphics[width=\figurewidth]{./OLS-under.pdf}
    \caption{Ordinary least-squares (OLS) fits (continuous lines) to a set of example data points (black dots). Fits are shown for different values of the number of basis functions $p$; there are $n=23$ data points. Here (and in all the \figurename s to follow) we are using the Fourier basis functions in \eqref{eq:basis} and shown in \figurename~\ref{fig:basis}. The data $y_i$ were generated using a function that does not reside in the function space spanned by the basis. Here the $X_\ast$ matrices used in the predictions $\hat{Y}_\ast$ were generated from a fine grid of locations in the location coordinate $t$; the plots of the fine grid of predictions are the continuous lines. The fits with larger $p$ have more flexibility to fit the data than the fit with $p=3$, but the fit at the highest $p$ shows evidence of over-fitting.}
    \label{fig:ols1}
    \end{mdframed}
\end{figure}

@@

Although the OLS prediction \eqref{eq:OLS} is affine invariant with respect to $p$-dimensional transformations, it is \emph{not} affine invariant with respect to $n$-dimensional transformations, such as a re-weighting of the input data. Indeed, if we know weights or inverse variances for our data points, conceptually we can put them into a weight matrix $C^{-1}$ (written this way to emphasize that reweighting is usually inverse-variance weighting) and write
\begin{equation}\label{eq:wls}
    \hat{Y}_\ast = X_\ast\,(X^\top C^{-1}\,X)^{-1}\,X^\top C^{-1}\,Y
    ~,
\end{equation}
where the weight matrix $C^{-1}$ is $n\times n$ (and often diagonal in standard applications). The weight matrix $C^{-1}$ is sometimes called the information tensor (or information matrix) and its inverse $C$ is often called the covariance matrix or the noise variance tensor.

This form \eqref{eq:wls} of least squares is called \emph{weighted least squares} (WLS) because of the data weighting (not to be confused with \emph{feature} weighting, to appear below).
It is also called \emph{chi-squared fitting} because it optimizes the scalar objective commonly called chi-squared:
\begin{equation}
    \chi^2 = (Y - X\,\beta)^\top C^{-1}\,(Y - X\,\beta)
    ~.
\end{equation}
%(If that isn't obviously chi-squared to you, recall that $C^{-1}$ is often diagonal and has the inverses of the squares of the data uncertainties on that diagonal.)
In \figurename~\ref{fig:wls} a comparison of OLS and WLS is shown, for a case of non-trivial data weights, where the data weights are set to be the inverse squares of individual data-point uncertainties.
\begin{figure}[t]
    \begin{mdframed}
    \includegraphics[width=\figurewidth]{./WLS.pdf}
    \caption{Comparison of ordinary least-squares (OLS) and weighted least-squares (WLS) fits (continuous lines) to the example data (black dots). In order to illustrate the differences, we assigned non-trivial error bars to the data. The error bars are ignored in the OLS fit, but in the WLS fit, the weight matrix $C^{-1}$ is diagonal with the diagonal entries set to the inverses of the squares of those error bars. The WLS fit ``pays less attention to'' the points on the left with the largest error bars.}
    \label{fig:wls}
    \end{mdframed}
\end{figure}

We will say more about noisy data and the propagation of uncertainties below in \sectionname~\ref{sec:uncertainty}.
It is true that in realistic scenarios there are uncertainties not just in the data points $y_i$, but also often in the \emph{locations} $t_i$ of the data as well. It turns out that taking the latter into account is a much harder problem; we will discuss this briefly in \sectionname~\ref{sec:uncertainty}.

@@

The ridge brings with it a choice: How to set the hyper-parameter $\lambda$?
We generally recommend cross-validation, to be discussed below in \sectionname~\ref{sec:dd}; cross-validation gives good results but it produces a slightly biased estimate \citep{cv-bias}.
Ridge regression is not affine invariant in the sense that the standard OLS prediction \eqref{eq:OLS} is; that is, the effect of the regularization depends on the amplitudes and linear combinations of features placed in the feature matrix.
This will become important later when we consider feature weights below.
It is also the case that the regularization need not be proportional to the identity matrix: In principle any positive definite matrix $\Lambda$ could be used in place of $\lambda\,I$; this makes sense to consider when we have detailed prior beliefs about all the parameters $\beta_j$ or if those parameters are measured with different units (say).

Combining both the point weighting from WLS and the generalized ridge regression into a weighted ridge yields to estimates
\begin{equation}\label{eq:foo}
    \hat{Y}_\ast = X_\ast\,(X^\top C^{-1}\,X + \Lambda)^{-1}\,X^\top C^{-1}\,Y
    ~.
\end{equation}
This form has good properties for many real-world physics applications, where data-point error bars are often known, and functions are often expected to be smooth.
We'll discuss this form \eqref{eq:foo} more below in \sectionname~\ref{sec:fwols},
but briefly we can say here that it appears in Bayesian inference contexts where the data points $y_i$ are treated as having Gaussian noise associated with them (with known variance tensor or covariance matrix $C$) and there is a Gaussian prior on the parameter vector $\beta$ (with known variance $\Lambda^{-1}$).
We have discussed that model elsewhere (\citealt{products}).\footnote{In our previous discussion of this product of Gaussians, the notation differs. What's called $\Lambda^{-1}$ here is called $\Lambda$ in \citet{products}.}
It is also useful sometimes to think of the ``units'' or dimensions of the quantities in \eqref{eq:foo}: In the Bayesian setting, the units of $C$ would be the square of the units of $Y$ and the units of $\Lambda^{-1}$ would be the square of the units of the ratio $Y/X$ (the square of the units of $\beta$).

@@

We are taught folklore, that we can never fit for more parameters than we have data. That is, we can never work at $p>n$.
This isn't true!
Not only is it the case that we \emph{can} work at $p>n$, in many cases we \emph{should} work at $p>n$, and many real-world regressions do.
But it is true that the $p>n$ regime is indeed strange.
In the over-parameterized case, there are typically many choices for the parameters $\beta_j$ that will literally zero out the differences between the data $Y$ and the linear prediction $X\,\beta$.
The OLS solution, in this case, is defined (somewhat arbitrarily) to be the minimum-norm parameter vector $\beta$ that interpolates the data:
\begin{equation}\label{eq:opt2}
    \hat{\beta} = \arg\min_\beta \|\beta\|_2^2 ~~\mbox{subject to}~~ Y = X\,\beta
    ~.
\end{equation}
Technically this formulation depends on an additional assumption that the feature matrix $X$ is full rank or that the data $Y$ lie in the subspace spanned by $X$.
This is true almost always when $p>n$.
This optimization is again convex and has a unique solution, although that solution will depend on feature weights that we discuss in a moment.
%When the investigator knows uncertainties on the training data points $y_i$, this expression will change a bit; we return to that in the next \sectionname.
The under-parameterized and over-parameterized optimization statements \eqref{eq:opt1} and \eqref{eq:opt2} can be unified into one form by considering the limit of light L2 regularization:
\begin{equation}\label{eq:opt3}
    \hat{\beta} = \lim_{\lambda\to 0^+}\left[\arg\min_\beta \|Y - X\,\beta\|_2^2 + \lambda\,\|\beta\|_2^2\right]
    ~;
\end{equation}
in the limit, this delivers the OLS solution in either case and doesn't require the constraint in \eqref{eq:opt2} to be satisfied.

In the over-parameterized case ($p>n$), the prediction looks like
\begin{equation}\label{eq:OLS2}
    \hat{Y}_\ast = X_\ast\,X^\top (X\,X^\top)^{-1}\,Y
    ~,
\end{equation}
provided that $X\,X^\top$ is invertible (which will usually be the case).
Like \eqref{eq:OLS}, this prediction is \emph{also} called ordinary least squares (OLS), or sometimes ``min-norm least-squares'' to emphasize the point that it is making the minimum-norm choice of $\beta$ among many degenerate solutions.
Examples of OLS fits in the over-parameterized regime are shown in \figurename~\ref{fig:ols2}.
Note that the fits with different numbers of parameters $p$ lead to very different predictions, but they all go through the data exactly.
\begin{figure}[t]
    \begin{mdframed}
    \includegraphics[width=\figurewidth]{./OLS-over.pdf}
    \caption{Ordinary least-squares (OLS) or min-norm least-squares fits (continuous lines) to a set of example data points (black dots), but now for a few over-parameterized cases. There are $n=23$ data points. As in \figurename~\ref{fig:ols1}, the data points $y_i$ were generated using a function that does not precisely reside in the function space spanned by the basis. The three fits are very different, but they all go through all the data points exactly. As $p$ gets large, the fit function approaches $y=0$ almost everywhere; this is a consequence of Plancharel's Theorem and the use of the Fourier basis (see text).}
    \label{fig:ols2}
    \end{mdframed}
\end{figure}

It is slightly off-topic to notice that in \figurename~\ref{fig:ols2}, as $p$ gets very large, the OLS solution approaches $y=0$ everywhere that it can.
This behavior is a direct consequence of using the Fourier basis and Plancharel's theorem, which states that the Fourier transform is unitary (see for instance \citealt{folland2016course}).
This means that the integral over frequency of the square of the Fourier transform is equal to the integral over location of the square of the original function.
Thus the min-norm solution delivered by OLS, which chooses the interpolating function that minimizes the squares of the component amplitudes $\beta$ in the Fourier basis used to make \figurename~\ref{fig:ols2}, will choose the interpolating function $y(t)$ that minimizes the mean square of the value of the function in the location space.
It will try to stay as close to $y=0$ as possible.
That's probably not desirable in most applications! We will fix that problem in the next \sectionname.

@@

In particular, the CVMSE for the feature-weighted OLS fits is bad when $p\approx n$, and much better at $p\ll n$ and $p \gg n$.
This phenomenology is not particular to this problem.
It is extremely general.
There is an effect known in many kinds of regression called ``double descent'' or ``peaking phenomenon'' or ``jamming'' in which predictive accuracy becomes very poor when the number of free parameters $p$ comes close to the number of data points $n$ (\citealt{jain198239}, \citealt{belkin2019reconciling}, \citealt{spigler2018jamming}, \citealt{geiger2019jamming}). For linear models, the ``risk''---the out-of-sample prediction error---blows up when the model capacity just becomes excessive at $p=n$ (\citealt{hastie2019surprises}).
The fundamental reason for this phenomenon is that the ordinary least-squares estimates \eqref{eq:OLS} and \eqref{eq:OLS2} require computation of the inverses $(X^\top X)^{-1}$ and $(X\,X^\top)^{-1}$ respectively, which are very badly conditioned around $p\approx n$.
This translates to a large variance for the estimator $\hat \beta$ which implies a large risk.
One way to think about this is that when the condition number of the matrix $X^\top X$ or $X\,X^\top$ is large, some directions in the data space---some linear combinations of the elements of the data vector $Y$---are very strongly amplified.\footnote{The condition number comes back up below in \sectionname~\ref{sec:implementation}. It turns out that when the condition number is large, not only do the output predictions become very sensitive to the input training data, but also the numerical (computational) stability of the linear algebra can also be badly affected. But we emphasize here that the risk goes bad when the condition number is large \emph{even if} it is possible to perform the linear algebra correctly at high precision.}
This makes the regression unreasonably sensitive to noise in the data.
The high-risk behavior at $p\approx n$ disappears with certain kinds of regularization.
These include the ridge regularization shown in \figurename~\ref{fig:cv} (also known as Tikhonov regression), but also early stopping (\citealt{hastie2019surprises}), and dimensionality reduction (\citealt{huang2020dimensionality}).
